---
title: New summarize_variable function
author: Kyle
date: '2020-08-14'
slug: demo-summarize-variable
categories:
  - Programming
tags:
  - R
cover: /img/post_cover/summarize_variable.png
---

I've made a new function (within [{kowr}](https://github.com/KoderKow/kowr)) for quickly summarizing variables in a dataset. I think `{skimr}` does a better job of summarizring the whole dataset. However, I find it hard to pipe the `{skimr}` results into a visualization. These outputs have helped me build quick and easy visualizations while undergoing in-depth EDA's (Exploratory Data Analysis).

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
```

# Libraries

The  `penguins` data is from the `{palmerpenguins}` package. I also use `{dplyr}`, `{ggplot2}`, `{tidyr}` and my personal `{kowr}` package for this demo.

```{r, message=FALSE, warning=FALSE}
library(palmerpenguins)
library(dplyr)
library(ggplot2)
library(tidyr)
library(kowr)
```

# Data

```{r}
penguins
```

# Summarize Character Data

```{r}
d <-
  penguins %>%
  summarize_variable(species)

d
```

The `summarize_variable()` function will arrange the character variable as a factor, ordered by count of the element.

```{r}
d %>% 
  ggplot() +
  aes(
    x = species,
    y = n,
    fill = species,
    label = p_format
  ) +
  geom_col(show.legend = FALSE) +
  geom_label(
    fill = "white",
    show.legend = FALSE
  ) + 
  scale_fill_kow() +
  theme_kow()
```

This will also work with `{dplyr}`'s `group_by()`.

```{r}
d <-
  penguins %>%
  group_by(island) %>%
  summarize_variable(species) %>% 
  arrange(island, species)

d
```

```{r}
d %>% 
  ggplot() +
  aes(
    x = island,
    y = n,
    fill = species,
    label = p_format
  ) +
  geom_col(position = "dodge") +
  geom_label(
    fill = "white",
    show.legend = FALSE
  ) + 
  scale_fill_kow() +
  theme_kow()
```

# Summarize Numeric Data

```{r}
d <-
  penguins %>%
  summarize_variable(bill_length_mm)

d
```

We can use these to plot lines on a histogram.

```{r}
d2 <-
  d %>%
  select(-c(n, sd, na_count)) %>% 
  pivot_longer(everything()) %>% 
  mutate(name = factor(
    x = name,
    levels = c("min", "q1", "mean", "median", "q3", "max")
  ))

d2
```

```{r}
penguins %>% 
  ggplot() +
  aes(x = bill_length_mm) +
  geom_histogram(alpha = 0.8) +
  geom_vline(
    data = d2,
    mapping = aes(
      xintercept = value,
      color = name
    )
  ) +
  scale_color_kow() +
  theme_kow()
```

```{r}
d <-
  penguins %>%
  group_by(species) %>%
  summarize_variable(bill_length_mm)

d
```

```{r}
d2 <-
  d %>%
  select(-c(n, sd, na_count)) %>% 
  pivot_longer(-species) %>% 
  mutate(name = factor(
    x = name,
    levels = c("min", "q1", "mean", "median", "q3", "max")
  ))

d2
```

```{r}
penguins %>% 
  ggplot() +
  aes(x = bill_length_mm) +
  geom_histogram(alpha = 0.8) +
  geom_vline(
    data = d2,
    mapping = aes(
      xintercept = value,
      color = name,
      group = species
    )
  ) +
  scale_color_kow() +
  theme_kow() +
  facet_wrap(
    facets = ~ species,
    nrow = 3
  )
```

